---
title: "LuminolData"
author: "Cameron Meikle"
date: "2024-10-13"
output:
  pdf_document: default
  html_document: default
---

Library packages and set the knitr
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Change working directory here
knitr::opts_chunk$set(root.dir = "/Users/cameronmeikle/Desktop/UNR_RA_position/Projects/Konzo")
#Checking that the repo is working


library(knitr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
library(ggpubr)

wd <- getwd()
```


Download and clean the datafile

13Oct2024 - Data is currently not manipulated in any way except for getting rid of the Dave samples 

```{r Datafile cleanup, echo=FALSE}
#Luminol Data
luminol_data <- read.csv(paste(wd,"/Luminol_field_data.csv", sep = ""))
#SubjectID_Data
Subject_ID_Data <- read.csv(paste(wd, "/KBS EDTA tube records_v1.csv", sep = ""))

#Clean the Subject_ID_Data and combine to dataframes
Subject_ID_Data <- Subject_ID_Data %>% rename(Subject_Barcode = Subject.Barcode, 
                                              Sample_Type = Sample.Type) %>% select(Subject_Barcode,Sample_Type) %>% filter(Sample_Type != "Cassava Pit H2O") %>%
  filter(Sample_Type != "DRM CONTROL")

```


```{r Datafile cleanup2, echo=FALSE}
#Resolve duplication problem
Subject_ID_Data <- Subject_ID_Data %>% distinct(Subject_Barcode, .keep_all = TRUE)

#gives you the mean value between the two replicates for the luminol data
luminol_data2 <- luminol_data %>% 
  group_by(Subject.ID) %>% 
  summarize(
    `10` = mean(`X10.minutes`, na.rm = TRUE),
    `12.5` = mean(`X12.5.minutes`, na.rm = TRUE), 
    `15` = mean(`X15.minutes`, na.rm = TRUE),
    across(c(Location, Date), first)
  )
#Change naming conventions of data
luminol_data <- luminol_data %>% rename( "15" = X15.minutes, "12.5" = X12.5.minutes,  "10" = X10.minutes)

#Filter out Dave Data
clean_luminol_data <- luminol_data2 %>% filter(Subject.ID != "Dave")

#Add Subject IDs
clean_luminol_data <- merge(clean_luminol_data, Subject_ID_Data, by.x = "Subject.ID", by.y = "Subject_Barcode")


Dave_Controls <- luminol_data %>% filter(Subject.ID == "Dave") %>% select(Subject.ID, Location, Date, Page, '10', '12.5', '15')

kable(Dave_Controls, caption = "Dave Controls")

```


Create basic histogram plots of the data
```{r overall histogram, echo=FALSE, warning=FALSE}
data_by_timepoint <- clean_luminol_data %>% pivot_longer( cols = c("10", "15", "12.5"), names_to = "Timepoint", values_to = "RFU")

#Overall histogram of values

ggplot(data_by_timepoint, aes(x = RFU, fill = Timepoint)) +
  geom_histogram(bins = 100) +
  labs(title = "Overall Histogram") +
  theme_minimal()

#Histogram by timepoint

unique_timepoints <- (c("10", "12.5", "15"))
for (timepoint in unique_timepoints) {
  df <- data_by_timepoint %>% filter(Timepoint == timepoint)
  
  histogram_plot <- ggplot(df, aes(x = RFU)) +
    geom_histogram(bins = 50) +
    labs(title = paste("Timepoint", timepoint, "minutes")) +
    theme_minimal()
  
  print(histogram_plot)
}

```

Create basic box plots of the data split by timepoint

```{r timepoint box plot, echo=FALSE, warning=FALSE}

ggplot(data_by_timepoint, aes(x = Timepoint, y = RFU)) +
  geom_boxplot() +
  stat_compare_means(method = "anova") +
  theme_minimal()

```


Based on the above graph, I think it is okay to say that we will restrict to minute 15 for the long term analysis. I kept everything in loops that include all timepoints at the moment but that can easily be changed.

I then break down the spread by location. Theoretically all locations should look similar. However, there is 
```{r, warnings=FALSE, message=FALSE, echo=FALSE}

for (timepoint in unique_timepoints) {
  df <- data_by_timepoint %>% filter(Timepoint == timepoint)
  
  box_plot <- ggplot(df, aes(x = Location, y = RFU)) +
  geom_boxplot() +
    labs(title = paste("Timepoint", timepoint, "Minutes")) +
    stat_compare_means(method = "anova") +
    theme_minimal()
  
  print(box_plot)
  
  anova <- aov(RFU ~ Location, df)
  
  print(summary(anova))
  
}

SamplesPerLocation <- clean_luminol_data %>% count(Location)
kable(SamplesPerLocation, caption = "Samples per location")
```


Boxplot comparing the controls versus diseased

```{r, warnings=FALSE, message=FALSE, echo=FALSE}
#Overall graphic
ggplot(data_by_timepoint, aes(x = Sample_Type, y = RFU)) +
  geom_boxplot() +
  stat_compare_means(method = "t.test") +
  theme_minimal()

#Split by timepoint
for (timepoint in unique_timepoints) {
  df <- data_by_timepoint %>% filter(Timepoint == timepoint)
  
  box_plot <- ggplot(df, aes(x = Sample_Type, y = RFU)) +
  geom_boxplot() +
  labs(title = paste("Timepoint", timepoint, "Minutes")) +
  stat_compare_means(method = "t.test") +
    theme_minimal()

  
  print(box_plot)
}

SamplesPerStatus <- clean_luminol_data %>% count(Sample_Type)
kable(SamplesPerStatus, caption = "Samples per Disease Status")

```

Since the above is significant, we are going to do the analysis on the 15 minute timepoint and split by location









```{r, echo=FALSE, warning=FALSE, message=FALSE}
unique_locations <- unique(data_by_timepoint$Location)

for (location in unique_locations) {
  df <- data_by_timepoint %>% filter(Timepoint == "15") %>% filter(Location == location)
  
  box_plot <- ggplot(df, aes(x = Sample_Type, y = RFU)) +
  geom_boxplot() +
  labs(title = paste("15 Minute Timepoint at", location)) +
  stat_compare_means(method = "t.test") +
    theme_minimal()
  
  print(box_plot)
  
  samples_df <- clean_luminol_data %>% filter(Location == location)
  SamplesPerStatus <- samples_df %>% count(Sample_Type)
  print(kable(SamplesPerStatus, caption = paste("Samples per Disease Status at", location)))
  
}

```









Future things to do:
-Figure out how to print the KNITR better (one page per graph with information)
-Look at differences between controls between villages (ANOVA) and then possibly a post-hoc (ttest)
-Think about possible "batch effect" problems

Once we have the correct data
-Check age as confounder
-Do analysis on stage
-Do analysis on recency



